---
title: "ENVS 193DS Final"
author: "Luma Lazarini"
date: "06/5/25"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[Github repo link](https://github.com/lumalazarini/ENVS-193DS_spring-2025_final)

# Set up

```{r, warning=FALSE, message=FALSE}
library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models
library(gt)
library(dplyr)
library(tidyr)
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(DHARMa)

```

# 1. Research writing

## a. Transparent statistical methods

In part 1, they used a Pearson's correlation test, since the data is parametric. In part 2, they used an ANOVA test.

## b. More information needed

One additional test my coworker should run is Tukey’s Honest Significant Difference (HSD) test, which allows for pairwise comparisons between nitrogen sources. This would show which specific sources (e.g., fertilizer vs. grasslands) have significantly different nitrogen loads, rather than just indicating that a difference exists overall.

In addition, they should report descriptive and test statistics such as the F-statistic, alpha value, and standard deviations for each source group. These values provide important context about the strength of the overall effect, the threshold for significance, and the variability within each source category, helping readers interpret the results more fully.

## c. Suggestions for rewriting

We found a statistically significant difference in mean annual nitrogen load among the five nitrogen sources (one-way ANOVA, F = \[F-statistic\], df = \[degrees of freedom\], p = 0.02, α = \[significance level\]).

Post-hoc Tukey HSD tests indicated that fertilizer sources contributed significantly more nitrogen than grasslands and atmospheric deposition (e.g., fertilizer vs. grasslands: difference = \[mean diff\], 95% CI: \[lower, upper\] kg year⁻¹).

On average, fertilizer sources had the highest nitrogen load (\[mean ± CI\] kg year⁻¹), while grasslands had the lowest. The effect size (η² = \[effect size\]) suggests a \[small/moderate/large\] impact of source type on nitrogen load.

# 2. Data visualization

```{r}
#reading in data
sst <- read_csv(here("data", "SST_update2023.csv"))
```

## a. Cleaning and summarizing

```{r}
#cleaning data
sst_clean <- sst |>
  clean_names() |> #clean column names
  mutate(
    year = year(date), #extract year from date in new column
    month = month(date, label = TRUE, abbr = TRUE)  # extract month in new column
  ) |> 
  # Group by year and month
  group_by(year, month) |> 
   summarise(
    mean_monthly_sst = mean(temp, na.rm = TRUE), .groups = "drop") |>   # <- replace `sst_c` if needed
  mutate(
    year = as.factor(year),
    month = factor(month, levels = month.abb, ordered = TRUE))


```

```{r}
# Show 5 random rows
slice_sample(sst_clean, n = 5)

```

```{r}
# Show structure
str(sst_clean)
  
```

## b. Visualize the data

```{r}
# Filter to only include 2018–2023
sst_subset <- sst_clean |> 
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023"))

# Create a color gradient from light to dark
ggplot(sst_subset, 
       aes(x = month, 
           y = mean_monthly_sst, 
           group = year, 
           color = year)) +
  geom_line(size = 1) + # first geometry
  geom_point(size = 2) + # second geometry
  scale_color_manual(values = scales::seq_gradient_pal("lightblue", "black", "Lab")
                     (seq(0, 1, length.out = 6))) +
  labs(
    x = "Month",
    y = "Mean monthly sea surface temperature (°C)",
    color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.1, .7), # legend inside the panel
    panel.border = element_rect(color = "black", fill = NA), # panel border
    panel.background = element_rect(fill = "white"), # white background
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank() #removing grid lines
  )

```

# 3. Data analysis

```{r}
#reading in data
nest_boxes <- read_csv(here("data", "occdist.csv")) |> 
  clean_names() #cleaning column names


```

## a. Response variable

In this dataset, 1s mean that a swift parrot was the species in the nest boxes, while 0s mean that another species was occupying the nest boxes.

## b. Purpose of study

Swift parrots are the target species, which are at critically risk of extinction. The other two species, the Common Starlings and Tree Martins, are non-target species that may compete with Swift Parrots for nesting sites.

## c. Difference in “seasons”

These two years (2016 and 2019) are referred to as seasons because they represent Swift Parrot breeding years triggered by mast flowering events. They are different because in 2016, nest boxes were newly installed, while in 2019 the same boxes were reused to compare nest box occupancy across time and assess whether permanent nest boxes benefit Swift Parrots or attract non-target species.

## d. Table of models

| Model number | Season | Distance to f/e | Description                |
|:------------:|:------:|:---------------:|:---------------------------|
|      0       |   No   |       No        | Null model                 |
|      1       |  Yes   |       Yes       | Saturated model            |
|      2       |  Yes   |       No        | only season as predictor   |
|      3       |   No   |       Yes       | only distance as predictor |

## e. Run the models

```{r model-fitting}
## Model fitting

# model 0: null model
model0 <- lm(
  sp ~ 1, # formula
  data = nest_boxes,# data frame
)

# model 1: all predictors (saturated model)
model1 <- lm(
  sp ~ season + edge_distance, # formula
  data = nest_boxes,# data frame
)

# model 2: season only
model2 <- lm(
  sp ~ season, # formula
  data = nest_boxes,# data frame
)

# model 3: distance only
model3 <- lm(
  sp~ edge_distance,# formula
  data = nest_boxes,# data frame
)
```

## f. Check the diagnostics

```{r model-diagnostics}
# Plot diagnostics
par(mfrow = c(2,2))
plot(model0)
plot(model1)
plot(model2)
plot(model3)
```


## g. Select the best model

```{r}
# Compare all models using AICc
AICc(model0,
     model1,
     model2,
     model3) |> 
  arrange(AICc)
```

```{r model-summary}
summary(model_)

```
```

